%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{}
\begin{center}
{\Large Introduction to Ancient Indian Mathamatics}
\end{center}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Introduction}
    
    \begin{itemize}
        \item \textbf{Numerals and Notation}: Contribution of decimal system and numerals like zero and place-value notation.
        \item \textbf{Sulba Sutras}: Texts dealing with geometric constructions for altars and other religious purposes.
        \item \textbf{Brahmagupta's Contributions}: Development of mathematical concepts including negative numbers and algebraic methods.
        \item \textbf{Aryabhata's Works}: Pioneering work in algebra, trigonometry, and approximation of pi.
        \item \textbf{Sanskrit Mathematical Literature}: Various treatises on arithmetic, geometry, and algebra.
    \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Introduction (Cont'd)}
    
    \begin{itemize}
        \item \textbf{Geometry}: Knowledge of geometric principles, including area and volume calculations.
        \item \textbf{Algebra}: Techniques for solving linear and quadratic equations.
        \item \textbf{Trigonometry}: Development of trigonometric functions and their applications in astronomy.
        \item \textbf{Number Theory}: Study of prime numbers, divisibility rules, and properties of integers.
        \item \textbf{Influence on Islamic Mathematics}: Indian mathematical concepts transmitted to the Islamic world, impacting further developments.
    \end{itemize}
\end{frame}

