%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{}
\begin{center}
{\Large Introduction to Ancient Indian Linguistics}
\end{center}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Introduction}
    
    \begin{itemize}
        \item \textbf{Panini's Grammar}: Panini's Ashtadhyayi is a foundational text in Sanskrit grammar.
        \item \textbf{Sanskrit Phonetics}: Detailed analysis of sounds, accents, and pronunciation.
        \item \textbf{Morphology and Syntax}: Rules for word formation and sentence structure.
        \item \textbf{Semantic Analysis}: Understanding the meaning and usage of words and phrases.
        \item \textbf{Vyakarana Tradition}: Development of grammatical theories and linguistic analysis.
    \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{ Introduction (Cont'd)}
    
    \begin{itemize}
        \item \textbf{Prakrit Languages}: Study of various regional languages and their grammatical structures.
        \item \textbf{Classifications}: Classification of languages based on their characteristics.
        \item \textbf{Etymology}: Analysis of word roots, origins, and evolution.
        \item \textbf{Alankara Shastra}: Study of figures of speech and literary ornamentation.
        \item \textbf{Influence on Literature}: Impact of linguistic theories on poetry and prose composition.
    \end{itemize}
\end{frame}
