%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Introduction to Yoganidra}
    \textbf{Yoga Nidra (योगनिद्रा)} is a deep relaxation technique that:
    \begin{itemize}
        \item Relieves stress.
        \item Improves sleep.
        \item Accesses the bliss state (Ananda आनन्द).
    \end{itemize}
    \textbf{Inspired by the Bihar School of Yoga}, this script follows the inward journey through the Koshas.
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Nidra vs Yoganidra}
    \textbf{Nidra (निद्रा)}: \\
    \begin{itemize}
        \item Unaware, only physical relaxation.
        \item Unconscious state.
    \end{itemize}
    \vspace{5mm}
    \textbf{Yoganidra (योगनिद्रा)}: \\
    \begin{itemize}
        \item Aware relaxation (physical, mental, and emotional).
        \item Conscious of subconscious mind.
    \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{8 Stages of Yoganidra}
    \begin{enumerate}
        \item \textbf{Preparation (Shavasana)}: Deep breaths in Shavasana (शवासन).
        \item \textbf{Resolve (Sankalpa)}: Optional positive affirmation (संकल्प).
        \item \textbf{Body Awareness (Rotation)}: Relax body parts.
        \item \textbf{Breath Awareness}: Relaxation through breath.
        \item \textbf{Opposite Sensations}: Experience and release emotions.
        \item \textbf{Visualization}: Reach the subconscious with imagery.
        \item \textbf{Resolve (Sankalpa)}: Repeat the Sankalpa again.
        \item \textbf{Exiting}: Return awareness to external surroundings.
    \end{enumerate}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Key Instructions}
    \begin{itemize}
        \item No movement during Yoganidra.
        \item Stay awake, do not fall asleep.
        \item Do not think, just follow the instructions.
    \end{itemize}
\end{frame}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{The Koshas (कोश)}
    \begin{itemize}
        \item \textbf{Annamaya Kosha (अन्नमयकोश)} - Physical Body
        \item \textbf{Pranamaya Kosha (प्राणमयकोश)} - Energy Body
        \item \textbf{Manomaya Kosha (मनोमयकोश)} - Emotional Body
        \item \textbf{Vijnanamaya Kosha (विज्ञानमयकोश)} - Wisdom Body
        \item \textbf{Anandamaya Kosha (आनन्दमयकोश)} - Bliss Body
    \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Koshas in Yoganidra}
    
    \begin{itemize}
        \item Body Awareness (Rotation): \textbf{Annamayakosha (अन्नमयकोश) - Physical Body:} Focus on different body parts (right palm, right arm, legs, back, etc.).
        \item Breath Awareness: \textbf{Pranamayakosha (प्राणमयकोश) - Breath Awareness:} Reverse breath count from 27.
        \item Opposite Sensations: \textbf{Manomayakosha (मनोमयकोश) - Emotional Body:} Experience opposite sensations (hot/cold, wet/dry).
        \item Visualization: \textbf{Vijnanamayakosha (विज्ञानमयकोश) - Subconscious Visualization:} Visualize calming scenes like deserts, lakes, and waves.
    \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Tips for Practicing Yoganidra}
    \begin{itemize}
        \item Use simple and precise language in the script.
        \item Speak in a clear and even tone.
        \item Sit comfortably and be still during facilitation.
        \item Practice in a warm, comfortable space. Use props (pillows, blankets) to support the body.
        \item Remain still, but do not fall asleep.
    \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Setting the Sankalpa (संकल्प)}
    \begin{itemize}
        \item A positive “I am” statement to guide your Yoganidra practice.
        \item Examples:
        \begin{itemize}
            \item "I am strong."
            \item "I am peaceful."
            \item "I am the witness."
        \end{itemize}
        \item Repeat the Sankalpa 3 times at the start and end of Yoganidra.
    \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Short Yoganidra Script}
    \textbf{Preliminary Relaxation}
    \begin{itemize}
        \item Lie in Shavasana (शवासन).
        \item Bring your awareness to the space between your body and the earth.
        \item Let your body soften and sink into the floor.
    \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Rotation of Awareness (Abbreviated)}
    \textbf{Focus on body parts:}
    \begin{multicols}{2}
    \begin{itemize}
        \item Right heel
        \item Left heel
        \item Right calf
        \item Left calf
        \item Right knee
        \item Left knee
        \item Right thigh
        \item Left thigh
    \end{itemize}
    \columnbreak
    \begin{itemize}
        \item Both hips
        \item Lower back
        \item Upper back
        \item Right shoulder
        \item Left shoulder
        \item Back of the head
    \end{itemize}
    \end{multicols}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Breath Awareness}
    \textbf{Breath Visualization:}
    \begin{itemize}
        \item Visualize breath as golden light flowing up and down the spine.
        \item Inhale: light rises from the tailbone to the crown.
        \item Exhale: light flows back down.
        \item Feel the cosmic flow of prana (प्राण).
    \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Opposite Sensations}
    \begin{itemize}
        \item Bring awareness to the sensation of heat.
        \item Feel your whole body becoming warm.
        \item Shift awareness to cold. Feel the entire body cooling down.
        \item Release both sensations.
    \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Guided Imagery}
    \textbf{Journey through Nature:}
    \begin{itemize}
        \item Imagine standing in a meadow, surrounded by a lush forest.
        \item Feel the warmth of the sun and smell the wildflowers.
        \item Walk into the forest, following a path that leads uphill.
        \item Reach a cave and discover a lit candle inside.
        \item Meditate on the candle's flame, with your Sankalpa inscribed on it.
    \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Exiting the Practice}
    \begin{itemize}
        \item Repeat your Sankalpa 3 times.
        \item Bring awareness to the sounds around you.
        \item Slowly move and break Shavasana.
    \end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Resources for Further Reading}
    \begin{itemize}
        \item \textbf{Books:}
        \begin{itemize}
            \item "Yoga Nidra" by Swami Satyananda Saraswati.
            \item "Yoga Nidra: A Meditative Practice for Deep Relaxation and Healing" by Richard Miller.
            \item "Yoga Nidra: The Art of Transformational Sleep" by Kamini Desai.
        \end{itemize}
    \end{itemize}
\end{frame}