%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{}
\begin{center}
{\Large Notes from Teachings of Dr Samparasd Vinod}
\end{center}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Meditation $dhyan$}

	\begin{itemize}
	\item Cannot observe thoughts as a third person
	\item Can not devoid of thoughts, force them to stop
	\item Just be friendly with them, seek why they are coming
	\item No need to sit for hours with pain in legs, just lie down.
	\end{itemize}

{\tiny (Ref:  आत्मयोग - परिपूर्ण जीवनशैली - YouTube)}

\end{frame}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Yoga}

	\begin{itemize}
	\item not obscure, out of realm but fun, easy but with clear crisp techniques
	\item though good for health, but its for spiritual upliftment
	\end{itemize}

{\tiny (Ref:  आत्मयोग - परिपूर्ण जीवनशैली - YouTube)}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Basic $dhyan$ Instructions}

	\begin{itemize}
	\item Sit comfortable either on chair or on ground
	\item Relax shoulders, neck, and keep back erect comfortably.
	\item Make sure you are very comfortable.
	\item Traverse from toe to head, see if there are any tensions, adjust them and release them.
	\item Experience how belly moves upon breathing, don't control, just experience.
	\item Let thoughts come, don't react, no judgment, just observe its journey.
	\end{itemize}

{\tiny (Ref:  ध्यान, प्राणायाम, ओंकार आणि प्रार्थना यांचा सराव - YouTube)}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Transition $dhyan$ Instructions}

	\begin{itemize}
	\item Breath going in and coming out, the changeover point is called 'sandhi', a transition
	\item Be aware of breath as well as Sandhi. Now keep watch on Sandhi.
	\end{itemize}

{\tiny (Ref:  ध्यान, प्राणायाम, ओंकार आणि प्रार्थना यांचा सराव - YouTube)}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Silent $dhyan$ Instructions}

	\begin{itemize}
	\item Breath going in and coming out, there should not be any sound, so, silent breathing.
	\item Inflate chest full, comfortably, deflate comfortably, slowly.
	\end{itemize}

{\tiny (Ref:  ध्यान, प्राणायाम, ओंकार आणि प्रार्थना यांचा सराव - YouTube)}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]\frametitle{Breathing Sound $dhyan$ Instructions}

	\begin{itemize}
	\item Breath going in and coming out, there should sound of it at throat, so, sound-full breathing.
	\item Inflate chest full, comfortably, deflate comfortably, slowly.
	\end{itemize}

{\tiny (Ref:  ध्यान, प्राणायाम, ओंकार आणि प्रार्थना यांचा सराव - YouTube)}

\end{frame}

